Fuel Economy and Analysis of Motor Trend Data Transmissions
========================================================

## Executive Summary

The contribution of automatic or manual transmissions (AM) to miles per gallon (MPG) was analyzed using the 1974 _Motor Trend Magazine_ data of 11 variables on 32 automobiles.  The resulting analysis of the _mtcars_ dataset in the standard R datasets package indicate:

- Cars with manual transmissions had better MPGs than automatic transmissions
- Holding all other variables constant, cars with manual transmissions had x.xxx MPG better fuel economy
- There were sevaral confounding variables including yyy, zzz which were included in the final model beyond transmission type
- The final model explained xxx of the variance as described by the $R^2$ value
- more
- more

## Data Transformation and Exploration

Several of the 11 variables in the dataset were initially numeric and had to be transformed into factor variables to be appropriately analyzed.

```{r echo=TRUE, results='hide'}
mtcars$am <- factor(mtcars$am, levels=c(0,1), labels=c("Automatic","Manual"))
mtcars$cyl <- factor(mtcars$cyl); mtcars$vs <- factor(mtcars$vs)
```
As can be seen in __Appendix Figure 1__, considering no other variables, there is a significant difference in MPG based on transmission type.

## Linear Models and Selection

A baseline model was initially developed using just trasmission type and MPG.
```{r baselm, highlight=TRUE, echo=TRUE}
(summaryBaselm <- summary(baselm <- lm(mpg ~ am, data=mtcars)))
(ciBaselm <- confint(baselm))  # confidence intervals of the coefficients
```
As can be seen, with no other variables there is a statistically significant difference of `r baselm$coefficients[2]` increase in MPG for manual transmissions (p=`r sprintf("%f", coef(summaryBaselm)[2,4])` , conf interval= (`r ciBaselm[2,1]` to `r ciBaselm[2,2]`)). However, the model only explains `r summary(baselm)$r.squared` of the variance in mpg based on the $R^2$ value.

## Conclusions


## Appendix
The following information is availabe as an appendix for review.

```{r fig.width=5, fig.height=5, echo=FALSE}
plot(mpg ~ am, mtcars, col="red",
     xlab="Transmission Type", ylab="Miles per Gallon",
     main="MPG by Transmission")
```
Figure 1: Automatic vs. Manual transmissions and Miles per Gallon with no other variables

```{r residuals, fig.width=7, fig.height=6, echo=FALSE}
par(mfrow=c(2,2))
plot(baselm)
```
Figure 2: Residuals analysis of base model including just mpg and transmission type


```{r}
summary(mtcars)
```