---
output: word_document
---
Fuel Economy and Analysis of Motor Trend Data Transmissions
========================================================

## Executive Summary

The contribution of automatic or manual transmissions (AM) to miles per gallon (MPG) was analyzed using the 1974 _Motor Trend Magazine_ data of 11 variables on 32 automobiles.  The resulting analysis of the _mtcars_ dataset in the standard R datasets package indicate:

- Cars with manual transmissions had better MPGs than automatic transmissions
- Holding all other variables constant, cars with manual transmissions had 7.25 MPG better fuel economy
- There were several confounding variables including weight in lbs/1000 and quarter mile time in seconds which were included in the final model in addition to transmission type
- The final model explained nearly 0.85 of the variance as described by the $R^2$ value
- In the final model including these additional variables, manual transmissions had better mileage but only about 2.9 MPG different than automatic transmission cars holding all other variables constant

## Data Transformation and Exploration

Several of the 11 variables in the dataset were initially numeric and had to be transformed into factor variables to be appropriately analyzed.

```{r echo=TRUE, results='hide'}
mtcars <- datasets::mtcars  # start with the baseline data from the datasets package
mtcars$am <- factor(mtcars$am, levels=c(0,1), labels=c("Automatic","Manual"))
mtcars$cyl <- factor(mtcars$cyl); mtcars$vs <- factor(mtcars$vs)
```
As can be seen in __Appendix Figure 1__, considering no other variables, there is a significant difference in MPG based on transmission type.

## Linear Models and Selection
__Base model:__
A baseline model was initially developed using just transmission type for an independent variable and MPG as the dependent.
```{r baselm, highlight=TRUE, echo=FALSE}
(summaryBaselm <- summary(baselm <- lm(mpg ~ am, data=mtcars)))
(ciBaselm <- confint(baselm))  # confidence intervals of the coefficients
```
As can be seen, with no other variables there is a statistically significant difference of `r baselm$coefficients[2]` increase in MPG for manual transmissions (p=`r sprintf("%f", coef(summaryBaselm)[2,4])`, conf interval= (`r ciBaselm[2,1]` to `r ciBaselm[2,2]`)). However, the model only explains `r summary(baselm)$r.squared` of the variance in mpg based on the $R^2$ value. (Residual tests of this model may be found in __Appendix Figure 3__.)

__Alternate model:__ However, in reviewing other parameters in the dataset, it was determined that including only the transmission type created a bias model.  We created several models incrementally adding parameters to find a possible optimum model with statistically significant parameters (P < 0.05) and strong $R^2$, variance inflation to find a parsimonious, interpretable representation.  Models were compared on these factors and using ANOVA tests.  The final model was then verified with the _step()_ function (see appendix for optional exploratory analysis) and compared to the singal variate model of mpg to transmission type using an ANOVA test.  We also looked at outliers using _hatvalues_ function, _dfbetas_ and _PRESS_ values on the most significant model to see if any of the model coefficients were significantly influence by specific outlier values.

```{r bestlmDisplay, echo=FALSE}
summaryBestlm <- summary(bestlm <- lm(mpg ~ wt + qsec + am, data=mtcars))
summaryBestlm 
anova (baselm, bestlm)
confint(bestlm)[2:4,]
```
It was determined that a model using transmission type, quarter mile time in seconds, and weight in lbs/1000 had a significantly better result (p-value for each coefficient < 0.05, $R^2$ = `r summaryBestlm$r.squared`) and was statistically significantly better than the base model based on anova tests (p = 1.6e-9). The confidence interval for the change in mpg for manual transmissions was reduced to `r coef(summaryBestlm)[4,1]` (conf interval = `r confint(bestlm)[4,1]` to `r confint(bestlm)[4,2]`) with these additional parameters added to the model.  The residual tests of this new model are available in __Appendix Figure 4__.

## Conclusions

Taken on its own, manual transmission cars would save on average `r coef(summaryBaselm)[2,1]` MPG if it was the only modeled independent variable.  However, a significantly better model with quarter mile time, weight and transmission type would explain about `r summaryBestlm$r.squared` of the variance in MPG.  Holding quarter mile time and weight constant, this model results in manual transmissions having `r coef(summaryBestlm)[4,1]` MPG better than automatic transmissions. 

In reviewing the residuals and PRESS values, the Datsun 710, Merc 230, Toyota Corona & Corolla, Fiat 128 and Chrysler Imperial had impacts on the regression (see __Appendix Figure 5__).

The reproducible research used to produce this report can be found at https://github.com/svonkleeck/RegressionProject.


# Appendix

The following information is available as an appendix for review or reference.

```{r fig.width=4, fig.height=4, echo=FALSE}
plot(mpg ~ am, mtcars, col="red",
     xlab="Transmission Type", ylab="Miles per Gallon",
     main="MPG by Transmission")
```

__Figure 1__: Automatic vs. Manual transmissions and Miles per Gallon with no other variables

```{r }
options(width=90)
summary(mtcars)
```

__Figure 2__: Summary of mtcars data used for analysis

```{r residualsBase, fig.width=6, fig.height=6, echo=FALSE}
par(mfrow=c(2,2), oma=c(0,0,1,0))
plot(baselm)#, main="Residual Analysis of Base Model")
```

__Figure 3__: Residuals analysis of base model including just mpg and transmission type


```{r residualsBest, fig.width=6, fig.height=6, echo=FALSE}
par(mfrow=c(2,2), oma=c(0,0,1,0))
plot(bestlm)
#mtext("Residual Analysis of Best Model", adj=0.5,  side=3, line=2, outer=TRUE)
```

__Figure 4__: Residuals analysis of best model including mpg, weight, quarter mile time and transmission type

```{r hatPRESS}
# influence potential
hat <- hatvalues(bestlm)
#hat[order(hat, decreasing = TRUE)]
PRESS<-resid(bestlm)/(1-hat)
PRESS[order(PRESS)]
dfbetas(bestlm)
```

__Figure 5__: Press values, dfbetas influence and hatvalues of influence on the best model

We can see the influence specific vehicles had on each coefficient.

```{r echo=FALSE, results='hide'}
# Some additional exploratory analysis for reproducible research but not included in the report
library(car)  # for vif() as per class

fitAll <- lm(mpg ~ ., data=mtcars)  # no good
summary(fitAll); vif(fitAll)

# Loop looking through many subsets of models for initial significant variables
# build model up from 1 am parameter
# this will help narrow parameters
newVars <- names(mtcars[!names(mtcars) %in% c("am", "mpg")]) # all names but am and mpg
baseformula <- "mpg ~ am"
prevlm <- baselm
# all formulas of length 2 with am
for (n  in newVars) {  #loop through all variables adding to base model
        formula <- paste(baseformula, n, sep="+")
        print(baseformula)
        newlm <- lm(formula, data=mtcars)
        print(summary(newlm))
        print(anova(newlm, baselm))
}

# Loop through adding variables
for (n  in newVars) {  #loop through all variables adding to base model
        baseformula <- paste(baseformula, n, sep="+")
        print(baseformula)
        newlm <- lm(baseformula, data=mtcars)
        print(summary(newlm))
        print(anova(newlm, baselm))
}

# As a result above, we keep seeing hp and wt
# Other significant methods explored:

# R2 78%, good p vals
(summaryBest2lm <- summary(best2lm <- lm(mpg ~ am + hp, data=mtcars))); 
anova (baselm, best2lm)

# This is good with R2 85%
(summaryBest2lm <- summary(best2lm <- lm(mpg ~ am + wt + qsec, data=mtcars))); 
anova (baselm, best2lm)
#Note AM and HP leaves AM signifcant and explains 78% variance but not best model as confirmed by step analysis.

# Optionally how to find the best fit using the step() function.  This is not part of the official analysis but was used to 
# verify the optimal model selection.
# Finally let's validated against car::step() best model


(summaryBestlmStep <- summary(bestlmStep <- step(lm(mpg ~ ., data=mtcars))))
```
