---
output: word_document
---
Fuel Economy and Analysis of Motor Trend Data Transmissions
========================================================

## Executive Summary

The contribution of automatic or manual transmissions (AM) to miles per gallon (MPG) was analyzed using the 1974 _Motor Trend Magazine_ data of 11 variables on 32 automobiles.  The resulting analysis of the _mtcars_ dataset in the standard R datasets package indicate:

- Cars with manual transmissions had better MPGs than automatic transmissions
- Holding all other variables constant, cars with manual transmissions had 7.25 MPG better fuel economy
- There were several confounding variables including weight in lbs/1000 and quarter mile time in seconds which were included in the final model in addition to transmission type
- The final model explained nearly 0.85 of the variance as described by the $R^2$ value
- In the final model including these additional variables, manual transmissions had better mileage but only about 2.9 MPG different than automatic transmission cars holding all other variables constant

## Data Transformation and Exploration

Several of the 11 variables in the dataset were initially numeric and had to be transformed into factor variables to be appropriately analyzed.

```{r echo=TRUE, results='hide'}
mtcars <- datasets::mtcars  # start with the baseline data from the datasets package
mtcars$am <- factor(mtcars$am, levels=c(0,1), labels=c("Automatic","Manual"))
mtcars$cyl <- factor(mtcars$cyl); mtcars$vs <- factor(mtcars$vs)
```
As can be seen in __Appendix Figure 1__, considering no other variables, there is a significant difference in MPG based on transmission type.

## Linear Models and Selection
__Base model:__
A baseline model was initially developed using just transmission type for an independent variable and MPG as the dependent.
```{r baselm, highlight=TRUE, echo=FALSE}
(summaryBaselm <- summary(baselm <- lm(mpg ~ am, data=mtcars)))
(ciBaselm <- confint(baselm))  # confidence intervals of the coefficients
```
As can be seen, with no other variables there is a statistically significant difference of `r baselm$coefficients[2]` increase in MPG for manual transmissions (p=`r sprintf("%f", coef(summaryBaselm)[2,4])`, conf interval= (`r ciBaselm[2,1]` to `r ciBaselm[2,2]`)). However, the model only explains `r summary(baselm)$r.squared` of the variance in mpg based on the $R^2$ value. (Residual tests of this model may be found in __Appendix Figure 3__.)

__Alternate model:__ However, in reviewing other parameters in the dataset, it was determined that including only the transmission type created a bias model.  We created several models using and then verified it with the _step()_ function (see appendix for optional exploratory analysis) and looked at the significance of comparing models using ANOVA tests, and the variance inflation to find a more parsimonious, interpretable representation.  We also looked at outliers using _hatvalues_ function, _dfbetas_ and _PRESS_ valueson the most significant model to see if any of the model coefficients were significantly influence by specific outlier values.


```{r bestlmDisplay, echo=FALSE}
summaryBestlm <- summary(bestlm <- lm(mpg ~ wt + qsec + am, data=mtcars))
summaryBestlm 
anova (baselm, bestlm)
```
It was determined that a model using transmission type, quarter mile time in seconds, and weight in lbs/1000 had a significantly better result (p-value for each coefficient < 0.05, $R^2$ = `r summaryBestlm$r.squared`) and was statistically significantly better than the base model based on anova tests (p = 1.6e-9). The confidence interval for manual transmissions was reduced to `r coef(summaryBestlm)[4,1]` (conf interval = `r confint(bestlm)[4,1]` to `r confint(bestlm)[4,2]`) with these additional parameters added to the model.  The residual tests of this new model are available in __Appendix Figure 4__.

## Conclusions

Taken on its own, manual transmission cars would save on average `r coef(summaryBaselm)[2,1]` MPG if it was the only modeled independent variable.  However, a significantly better model with quarter mile time, weight and transmission type would explain about `r summaryBestlm$r.squared` of the variance in MPG.  Holding quarter mile time and weight constant, this model results in manual transmissions having `r coef(summaryBestlm)[4,1]` MPG better than automatic transmissions. 

In reviewing the residuals and PRESS values, the Datsun 710, Merc 230, Toyota Corona & Corolla, Fiat 128 and Chrysler Imperial had impacts on the regression (see __Appendix Figure 5__).

The reproducible research used to produce this report can be found at https://github.com/svonkleeck/RegressionProject.


# Appendix

The following information is available as an appendix for review or reference.

```{r fig.width=4, fig.height=4, echo=FALSE}
plot(mpg ~ am, mtcars, col="red",
     xlab="Transmission Type", ylab="Miles per Gallon",
     main="MPG by Transmission")
```

__Figure 1__: Automatic vs. Manual transmissions and Miles per Gallon with no other variables

```{r }
options(width=90)
summary(mtcars)
```

__Figure 2__: Summary of mtcars data used for analysis

```{r residualsBase, fig.width=6, fig.height=6, echo=FALSE}
par(mfrow=c(2,2), oma=c(0,0,1,0))
plot(baselm)#, main="Residual Analysis of Base Model")
```

__Figure 3__: Residuals analysis of base model including just mpg and transmission type


```{r residualsBest, fig.width=6, fig.height=6, echo=FALSE}
par(mfrow=c(2,2), oma=c(0,0,1,0))
plot(bestlm)
#mtext("Residual Analysis of Best Model", adj=0.5,  side=3, line=2, outer=TRUE)
```

__Figure 4__: Residuals analysis of best model including mpg, weight, quarter mile time and transmission type

```{r hatPRESS}
# influence potential
hat <- hatvalues(bestlm)
#hat[order(hat, decreasing = TRUE)]
PRESS<-resid(bestlm)/(1-hat)
PRESS[order(PRESS)]
dfbetas(bestlm)
```

__Figure 5__: Press values, dfbetas influence and hatvalues of influence on the best model

We can see the influence specific vehicles had on each coefficient.

### Exploring aternate models
Optionally how to find the best fit using the step() function.  This is not part of the official analysis but was used to verify the optimal model selection.

```{r findBest, results='hide', echo=TRUE}
library(car)
(summaryBestlmStep <- summary(bestlmStep <- step(lm(mpg ~ ., data=mtcars))))
```
Other significant methods explored:
```
fitAll <- lm(mpg ~ ., data=mtcars)  # no good
summary(fitAll); vif(fitAll)

(summaryBest2lm <- summary(best2lm <- lm(mpg ~ am + hp, data=mtcars))); 
anova (baselm, best2lm)
```
Note AM and HP leaves AM signifcant and explains 78% variance but not best model as confirmed by step analysis.

